### 2018-05-02 doc/paper.tex
Adding the content of the "Numerical experiements" section.
Also adding the start of a theorem for the a priori error bound.

\renewcommand{\headrulewidth}{0pt}
\newcommand{\cond}[1]{\operatorname{cond}\left(#1\right)}
\newcommand{\fl}[1]{\operatorname{fl}\left(#1\right)}
\newcommand{\eps}{\varepsilon}
\newcommand{\mach}{\mathbf{u}}

@@

\begin{breakablealgorithm}
  \caption{\textit{de Casteljau algorithm for polynomial evaluation.}}

@@

  \begin{algorithmic}
    \Function{\(\mathtt{res} = \mathtt{CompDeCasteljau}\)}{$b, s$}
      \State \(n = \texttt{length}(b) - 1\)
      \State \(\left[\widehat{r}, \rho\right] = \mathtt{TwoSum}(1, -s)\)
      \\
      \For{\(j = 0, \ldots, n\)}
        \State \(\widehat{b}_j^{(n)} = b_j\)
        \For{\(F = 1, \ldots, K - 1\)}
          \State \(\widehat{\partial^F b}_j^{(n)} = 0\)
        \EndFor
      \EndFor
      \\
      \For{\(k = n - 1, \ldots, 0\)}
        \For{\(j = 0, \ldots, k\)}
          \State \(\left[P_1, \pi_1\right] = \mathtt{TwoProd}\left(
              \widehat{r}, \widehat{b}_j^{(k + 1)}\right)\)
          \State \(\left[P_2, \pi_2\right] = \mathtt{TwoProd}\left(
              s, \widehat{b}_{j + 1}^{(k + 1)}\right)\)
          \State \(\left[\widehat{b}_j^{(k)}, \sigma_3\right] =
              \mathtt{TwoSum}(P_1, P_2)\)
          \\
          \State \(e = \left[\pi_1, \pi_2, \sigma_3\right]\)
          \State \(\delta b = \widehat{b}_j^{(k + 1)}\)
          \\
          \For{\(F = 1, \ldots, K - 2\)}
            \State \(\left[\eta, \widehat{\ell}\right] =
                \mathtt{LocalErrorEFT}(e, \rho, \delta b)\)
            \State \(M = \texttt{length}(\eta)\)
            \\
            \State \(\left[P_1, \eta_{M + 1}\right] = \mathtt{TwoProd}\left(
                \widehat{r}, \widehat{\partial^F b}_j^{(k + 1)}\right)\)
            \State \(\left[P_2, \eta_{M + 2}\right] = \mathtt{TwoProd}\left(
                s, \widehat{\partial^F b}_{j + 1}^{(k + 1)}\right)\)
            \State \(\left[S_3, \eta_{M + 3}\right] =
                \mathtt{TwoSum}(P_1, P_2)\)
            \State \(\left[\widehat{\partial^F b}_j^{(k)}, \eta_{M + 4}\right]
                = \mathtt{TwoSum}\left(S_3, \widehat{\ell}\right)\)
            \\
            \State \(e = \eta\)
            \State \(\delta b = \widehat{\partial^F b}_j^{(k + 1)}\)
          \EndFor
          \\
          \State \(\widehat{\ell} =
                \mathtt{LocalError}(e, \rho, \delta b)\)
          \State \(\widehat{\partial^{K - 1} b}_j^{(k)} =
              \left(\widehat{r} \otimes
              \widehat{\partial^{K - 1} b}_j^{(k + 1)}\right) \oplus
              \left(s \otimes \widehat{\partial^{K - 1} b}_{j + 1}^{(k + 1)}
              \right) \oplus
              \widehat{\ell}\)
        \EndFor
      \EndFor
      \\
      \State \(\mathtt{res} = \widehat{b}_0^{(0)}\)
      \For{\(F = 1, \ldots, K - 1\)}
        \State \(\mathtt{res} = \mathtt{res} \oplus
            \widehat{\partial^F b}_0^{(0)}\)
      \EndFor
    \EndFunction
  \end{algorithmic}
\end{breakablealgorithm}

@@

@@

Figure~\ref{fig:smooth-drawing} and Figure~\ref{fig:compensated-k}
show how things improve.

@@

@@

>>>>

\renewcommand{\headrulewidth}{0pt}
\renewcommand{\qed}{\(\blacksquare\)}
\newcommand{\cond}[1]{\operatorname{cond}\left(#1\right)}
\newcommand{\fl}[1]{\operatorname{fl}\left(#1\right)}
\newcommand{\eps}{\varepsilon}
\newcommand{\mach}{\mathbf{u}}

@@

\begin{breakablealgorithm}
  \caption{\textit{de Casteljau algorithm for polynomial evaluation.}}
  \label{alg:de-casteljau}

@@

  \begin{algorithmic}
    \Function{\(\mathtt{res} = \mathtt{CompDeCasteljauK}\)}{$b, s, K$}
      \State \(n = \texttt{length}(b) - 1\)
      \State \(\left[\widehat{r}, \rho\right] = \mathtt{TwoSum}(1, -s)\)
      \\
      \For{\(j = 0, \ldots, n\)}
        \State \(\widehat{b}_j^{(n)} = b_j\)
        \For{\(F = 1, \ldots, K - 1\)}
          \State \(\widehat{\partial^F b}_j^{(n)} = 0\)
        \EndFor
      \EndFor
      \\
      \For{\(k = n - 1, \ldots, 0\)}
        \For{\(j = 0, \ldots, k\)}
          \State \(\left[P_1, \pi_1\right] = \mathtt{TwoProd}\left(
              \widehat{r}, \widehat{b}_j^{(k + 1)}\right)\)
          \State \(\left[P_2, \pi_2\right] = \mathtt{TwoProd}\left(
              s, \widehat{b}_{j + 1}^{(k + 1)}\right)\)
          \State \(\left[\widehat{b}_j^{(k)}, \sigma_3\right] =
              \mathtt{TwoSum}(P_1, P_2)\)
          \\
          \State \(e = \left[\pi_1, \pi_2, \sigma_3\right]\)
          \State \(\delta b = \widehat{b}_j^{(k + 1)}\)
          \\
          \For{\(F = 1, \ldots, K - 2\)}
            \State \(\left[\eta, \widehat{\ell}\right] =
                \mathtt{LocalErrorEFT}(e, \rho, \delta b)\)
            \State \(M = \texttt{length}(\eta)\)
            \\
            \State \(\left[P_1, \eta_{M + 1}\right] = \mathtt{TwoProd}\left(
                \widehat{r}, \widehat{\partial^F b}_j^{(k + 1)}\right)\)
            \State \(\left[P_2, \eta_{M + 2}\right] = \mathtt{TwoProd}\left(
                s, \widehat{\partial^F b}_{j + 1}^{(k + 1)}\right)\)
            \State \(\left[S_3, \eta_{M + 3}\right] =
                \mathtt{TwoSum}(P_1, P_2)\)
            \State \(\left[\widehat{\partial^F b}_j^{(k)}, \eta_{M + 4}\right]
                = \mathtt{TwoSum}\left(S_3, \widehat{\ell}\right)\)
            \\
            \State \(e = \eta\)
            \State \(\delta b = \widehat{\partial^F b}_j^{(k + 1)}\)
          \EndFor
          \\
          \State \(\widehat{\ell} =
                \mathtt{LocalError}(e, \rho, \delta b)\)
          \State \(\widehat{\partial^{K - 1} b}_j^{(k)} =
              \left(\widehat{r} \otimes
              \widehat{\partial^{K - 1} b}_j^{(k + 1)}\right) \oplus
              \left(s \otimes \widehat{\partial^{K - 1} b}_{j + 1}^{(k + 1)}
              \right) \oplus
              \widehat{\ell}\)
        \EndFor
      \EndFor
      \\
      \State \(\mathtt{res} = \widehat{b}_0^{(0)}\)
      \For{\(F = 1, \ldots, K - 1\)}
        \State \(\mathtt{res} = \mathtt{res} \oplus
            \widehat{\partial^F b}_0^{(0)}\)
      \EndFor
    \EndFunction
  \end{algorithmic}
\end{breakablealgorithm}

@@

\begin{theorem}
  If no underflow occurs, \(n \geq 2\) and \(s \in \left[0, 1\right]\)
  \begin{equation}
    \frac{\left|p(s) - \mathtt{CompDeCasteljau}(p, s, K)\right|}{
      \left|p(s)\right|} \leq \left[\mach + \mathcal{O}\left(\mach^2\right)
      \right] +
      \left[\left(\alpha(n) \mach\right)^K + \mathcal{O}\left(
      \mach^{K + 1}\right)\right] \cond{p, s}.
  \end{equation}
\end{theorem}

\begin{proof}
Note that
\begin{equation}
p(s) = b_0^{(0)} = \widehat{b}_0^{(0)} + \partial b_0^{(0)}
= \cdots
= \widehat{b}_0^{(0)} + \widehat{\partial b}_0^{(0)} + \cdots
+ \widehat{\partial^{K - 2} b}_0^{(0)} + \partial^{K - 1} b_0^{(0)}
\end{equation}
and
\begin{equation}
\mathtt{CompDeCasteljau}(p, s, K) = \widehat{b}_0^{(0)} \oplus
\widehat{\partial b}_0^{(0)} \oplus \cdots
\oplus \widehat{\partial^{K - 2} b}_0^{(0)} \oplus
\widehat{\partial^{K - 1} b}_0^{(0)}.
\end{equation}
It will be crucial for us to bound
\(\partial^K b_0^{(0)} = \partial^{K - 1} b_0^{(0)} -
\widehat{\partial^{K - 1} b}_0^{(0)}\). In addition, note that
\begin{align}
&\widehat{b}_0^{(0)} \oplus
\widehat{\partial b}_0^{(0)} \oplus \cdots
\oplus \widehat{\partial^{K - 2} b}_0^{(0)} \oplus
\widehat{\partial^{K - 1} b}_0^{(0)} \\
&= \widehat{b}_0^{(0)} \left(1 + \theta_{K - 1}\right) +
\widehat{\partial b}_0^{(0)} \left(1 + \theta_{K - 1}\right) + \cdots
+ \widehat{\partial^{K - 2} b}_0^{(0)} \left(1 + \theta_2\right) +
\widehat{\partial^{K - 1} b}_0^{(0)} \left(1 + \theta_1\right).
\end{align}
We'll use these two quantities to prove a bound.
\end{proof}

@@

All experiments were performed in IEEE-754 double precision.
As in \cite{Jiang2010}, we consider the evaluation in the neighborhood
of the multiple root of \(p(s) = (s - 1)\left(s - \frac{3}{4}\right)^7\),
written in Bernstein form.

@@

Figure~\ref{fig:smooth-drawing} shows the evaluation of \(p(s)\) at
the 401 equally spaced\footnote{It's worth noting that \(0.1\) cannot
be represented exactly in IEEE-754 double precision (or any binary
arithmetic for that matter). Hence (most of) the points of the form
\(a + b \cdot 10^{-c}\) can only be approximately represented.} points
\(\left\{\frac{3}{4} + j \frac{10^{-7}}{2}\right\}_{j=-200}^{200}\)
with \texttt{DeCasteljau} (Algorithm~\ref{alg:de-casteljau}),
\texttt{CompDeCasteljau} (Algorithm~\ref{alg:comp-de-casteljau})
and \texttt{CompDeCasteljau3} (Algorithm~\ref{alg:k-comp-de-casteljau}
with \(K = 3\)). We see that \texttt{DeCasteljau} fails to get the
magnitude correct, \texttt{CompDeCasteljau} has the right shape but
lots of noise and \texttt{CompDeCasteljau3} is able to smoothly evaluate
the function. This is in contrast to a similar figure in \cite{Jiang2010},
where the plot was smooth for the 400 equally spaced points
\(\left\{\frac{3}{4} + \frac{10^{-4}}{2} \frac{2j - 399}{399}
\right\}_{j=0}^{399}\). The primary difference is that as the interval
shrinks by a factor of \(\approx \frac{10^{-4}}{10^{-7}} = 10^3\), the
condition number goes up by \(\approx 10^{21}\) and \texttt{CompDeCasteljau}
is no longer accurate.

@@

Figure~\ref{fig:compensated-k} shows the relative forward errors compared
against the condition number. To compute relative errors, each input and
coefficient is converted to a fraction (i.e. infinite precision) and then
compared to the corresponding computed values. Similar tools are used to
\textbf{exactly} compute the condition number, though here we can rely
on the fact that \(\widetilde{p}\left(s\right) = (s - 1)
\left(\frac{s}{2} - \frac{3}{4}\right)^7\). Once the relative errors and
condition numbers are computed as fractions, they are rounded to the
nearest IEEE-754 double precision value. As in \cite{Jiang2010}, we use
values \(\left\{\frac{3}{4} - (1.3)^j\right\}_{j=-5}^{-90}\)\footnote{As with
\(0.1\), it's worth noting that \((1.3)^j\) can't be represented exactly in
IEEE-754 double precision. However, this geometric series still serves a
useful purpose since it continues to raise \(\cond{p, s}\) as \(j\) decreases
away from \(0\) and because it results in ``random'' changes in the bits of
\(0.75\) that are impacted by subtracting \((1.3)^j\).}. The curves for
\texttt{DeCasteljau} and \texttt{CompDeCasteljau} trace the same paths seen
in \cite{Jiang2010}. In particular, \texttt{CompDeCasteljau} has a relative
error that is \(\mathcal{O}(\mach)\) until \(\cond{p, s}\) reaches
\(1 / \mach\), at which point the relative error increases linearly with
the condition number until it becomes \(\mathcal{O}(1)\) when
\(\cond{p, s}\) reaches \(1 / \mach^2\).
Similarly, the relative error in \texttt{CompDeCasteljau3}
(Algorithm~\ref{alg:k-comp-de-casteljau} with \(K = 3\))
is \(\mathcal{O}(\mach)\) until \(\cond{p, s}\) reaches
\(1 / \mach^2\) at which point the relative error increases linearly
to \(\mathcal{O}(1)\) when \(\cond{p, s}\) reaches \(1 / \mach^3\)
and the relative error in \texttt{CompDeCasteljau4}
(Algorithm~\ref{alg:k-comp-de-casteljau} with \(K = 4\))
is \(\mathcal{O}(\mach)\) until \(\cond{p, s}\) reaches
\(1 / \mach^3\) at which point the relative error increases linearly
to \(\mathcal{O}(1)\) when \(\cond{p, s}\) reaches \(1 / \mach^4\).
